---
title: "STAT 429 Group Project - Multiple Linear Regression of WHO Life Expectancy
  Data"
author:
- "Jina Park - 30129680"
- "Scott Holden - 30051473"
- "Leigh Murray - 30075588"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


library(readxl)
library(car)
library(MASS)

```


```{r}
Life_Expectancy_Data_raw <- read_excel("C:/Users/User/Desktop/STAT 429/Life_Expectancy_Data.xlsx")

attach(Life_Expectancy_Data_raw)
names(Life_Expectancy_Data_raw)

dim(Life_Expectancy_Data_raw)
#[1] 2938   22

sum(is.na(Life_Expectancy_Data_raw)) # High presence of NULL values
```

Here we remove the rows containing null values.

```{r}
Life_Expectancy_Data<- na.omit(Life_Expectancy_Data_raw)
remove(Life_Expectancy_Data_raw)
dim(Life_Expectancy_Data)
#[1] 1649   22

length(Life_Expectancy_Data) #variables

#how many mixing values we have for each variable

sum(is.na(Life_Expectancy_Data))
#[1] 0
```

# Variables and Descriptions

Country - Country Name

Year - Year of the data

Status - Country Status of developed or developing

Life_Expectancy - Life Expectancy in Age (Response)

Adult_Mortality - Adult mortality rates (Probability of dying between 15 and 60 years per 1000 population)

infant.deaths - Number of infant deaths per 1000 population

Alcohol - Alcohol consumption per capita

Percentage Expenditure - Expenditure on health as percentage of GDP

Hepatitis B - HepB immunization coverage

Measles - Number of reported cases per 1000 population

BMI - Average Body mass index of entire population

under_five_deaths - number of under five deaths per 1000 population

Polio - Polio Immunization coverage among 1 year olds

Total Expenditure - General government expenditure on health as percentage of total government expenditure

Diphtheria - DTP3 Immunization coverage among 1 year olds

HIV.AIDS - Deaths per 1000 live births with HIV/ AIDS (0-4 years)

GDP - Gross Domestic product per capita

Population - Population of the country

thinness_10-19 years - prevelance of thinness among children aged 10-19

thinness_5-9 years - prevelance of thinness among children aged 5 - 9

Income Composition of Resources - Human development index

Schooling - Number of years of schooling 


#### Country and Status are both categorical variables, additionally Year is not considered as we seek to predict life expectancy regardless or time or place. Additionally, infant deaths is captured by under 5 deaths. So we can remove this variable also. 

```{r}
Life_Expectancy_Data <- subset(Life_Expectancy_Data, select = -c(Country, Year, Status, `infant deaths`))
```

# Cleaning and EDA

```{r}
head(Life_Expectancy_Data[0:6])
head(Life_Expectancy_Data[7:12])
head(Life_Expectancy_Data[13:18])
```


```{r}
summary(Life_Expectancy_Data$`Life expectancy`)

summary(Life_Expectancy_Data$`Adult Mortality`)

summary(Life_Expectancy_Data$Alcohol)

summary(Life_Expectancy_Data$`percentage expenditure`)

summary(Life_Expectancy_Data$`Hepatitis B`)

summary(Life_Expectancy_Data$Measles)

summary(Life_Expectancy_Data$BMI)

summary(Life_Expectancy_Data$Polio)

summary(Life_Expectancy_Data$`under-five deaths`)

summary(Life_Expectancy_Data$`Total expenditure`)

summary(Life_Expectancy_Data$`Diphtheria`)

summary(Life_Expectancy_Data$`HIV/AIDS`)

summary(Life_Expectancy_Data$GDP)

summary(Life_Expectancy_Data$Population)

summary(Life_Expectancy_Data$`thinness  1-19 years`)

summary(Life_Expectancy_Data$`thinness 5-9 years`)

summary(Life_Expectancy_Data$`Income composition of resources`)

summary(Life_Expectancy_Data$Schooling)
```
## Visual Inspection of Data
```{r}

par(mfrow=c(1,4))
boxplot(`percentage expenditure`,main="Percentage Expenditure")
boxplot(Measles,main="Measles")
boxplot(`under-five deaths`,main="under-five deaths")
boxplot(`HIV/AIDS`,main="HIV/AIDS")

par(mfrow=c(1,5))
boxplot(`Adult Mortality`, main= "Adult Mortality")
boxplot(Alcohol, main="Alcohol Cons.")
boxplot(`Hepatitis B`, main="Hepatitis B")
boxplot(BMI, main = "BMI")
boxplot(Polio, main= "Polio")

par(mfrow=c(1,5))
boxplot(`Total expenditure`, main="Total Expenditure")
boxplot(Diphtheria, main="Diphtheria")
boxplot(Measles, main="Measles")
boxplot(Population, main="Population")
boxplot(`thinness  1-19 years`, main="Thinness in children 10-19 years old")

par(mfrow=c(1,3))
boxplot(`thinness 5-9 years`, main="Thinness in children 5-9 years old")
boxplot(`Income composition of resources`, main="Income Composition of Resources")
boxplot(Schooling, main="Schooling")

```

The boxplots indicate that the majority of the predictor variables have lots of outliers. Only Alcohol consumption, BMI, Income Composition of Resources and Schooling contain few or no outliers.

```{r}
#This one should be our y-variables
par(mfrow=c(1,1))
boxplot(`Life expectancy`,main="Life expectancy",
col="mediumaquamarine")
```

```{r}
par(mfrow=c(3,2))

plot(Life_Expectancy_Data$`Adult Mortality`, Life_Expectancy_Data$`Life expectancy`)
abline(lsfit(Life_Expectancy_Data$`Adult Mortality`, Life_Expectancy_Data$`Life expectancy`), lty=2, col='red', lwd=2)

plot(Life_Expectancy_Data$Alcohol, Life_Expectancy_Data$`Life expectancy`)
abline(lsfit(Life_Expectancy_Data$Alcohol, Life_Expectancy_Data$`Life expectancy`), lty=2, col='red', lwd=2)

plot(Life_Expectancy_Data$`percentage expenditure`, Life_Expectancy_Data$`Life expectancy`)
abline(lsfit(Life_Expectancy_Data$`percentage expenditure`, Life_Expectancy_Data$`Life expectancy`), lty=2, col='red', lwd=2)

plot(Life_Expectancy_Data$`Hepatitis B`, Life_Expectancy_Data$`Life expectancy`)
abline(lsfit(Life_Expectancy_Data$`Hepatitis B`, Life_Expectancy_Data$`Life expectancy`), lty=2, col='red', lwd=2)


plot(Life_Expectancy_Data$Measles, Life_Expectancy_Data$`Life expectancy`)
abline(lsfit(Life_Expectancy_Data$Measles, Life_Expectancy_Data$`Life expectancy`), lty=2, col='red', lwd=2)

```

```{r}
par(mfrow=c(3,2))

plot(Life_Expectancy_Data$BMI, Life_Expectancy_Data$`Life expectancy`)
abline(lsfit(Life_Expectancy_Data$BMI, Life_Expectancy_Data$`Life expectancy`), lty=2, col='red', lwd=2)

plot(Life_Expectancy_Data$Polio, Life_Expectancy_Data$`Life expectancy`)
abline(lsfit(Life_Expectancy_Data$Polio, Life_Expectancy_Data$`Life expectancy`), lty=2, col='red', lwd=2)

plot(Life_Expectancy_Data$`under-five deaths`, Life_Expectancy_Data$`Life expectancy`)
abline(lsfit(Life_Expectancy_Data$`under-five deaths`, Life_Expectancy_Data$`Life expectancy`), lty=2, col='red', lwd=2)

plot(Life_Expectancy_Data$`Total expenditure`, Life_Expectancy_Data$`Life expectancy`)
abline(lsfit(Life_Expectancy_Data$`Total expenditure`, Life_Expectancy_Data$`Life expectancy`), lty=2, col='red', lwd=2)


plot(Life_Expectancy_Data$`Diphtheria`, Life_Expectancy_Data$`Life expectancy`)
abline(lsfit(Life_Expectancy_Data$`Diphtheria`, Life_Expectancy_Data$`Life expectancy`), lty=2, col='red', lwd=2)

plot(Life_Expectancy_Data$`HIV/AIDS`, Life_Expectancy_Data$`Life expectancy`)
abline(lsfit(Life_Expectancy_Data$`HIV/AIDS`, Life_Expectancy_Data$`Life expectancy`), lty=2, col='red', lwd=2)
```

```{r}
par(mfrow=c(3,2))

plot(Life_Expectancy_Data$GDP, Life_Expectancy_Data$`Life expectancy`)
abline(lsfit(Life_Expectancy_Data$GDP, Life_Expectancy_Data$`Life expectancy`), lty=2, col='red', lwd=2)

plot(Life_Expectancy_Data$Population, Life_Expectancy_Data$`Life expectancy`)
abline(lsfit(Life_Expectancy_Data$Population, Life_Expectancy_Data$`Life expectancy`), lty=2, col='red', lwd=2)

plot(Life_Expectancy_Data$`thinness  1-19 years`, Life_Expectancy_Data$`Life expectancy`)
abline(lsfit(Life_Expectancy_Data$`thinness  1-19 years`, Life_Expectancy_Data$`Life expectancy`), lty=2, col='red', lwd=2)

plot(Life_Expectancy_Data$`thinness 5-9 years`, Life_Expectancy_Data$`Life expectancy`)
abline(lsfit(Life_Expectancy_Data$`thinness 5-9 years`, Life_Expectancy_Data$`Life expectancy`), lty=2, col='red', lwd=2)


plot(Life_Expectancy_Data$`Income composition of resources`, Life_Expectancy_Data$`Life expectancy`)
abline(lsfit(Life_Expectancy_Data$`Income composition of resources`, Life_Expectancy_Data$`Life expectancy`), lty=2, col='red', lwd=2)

plot(Life_Expectancy_Data$Schooling, Life_Expectancy_Data$`Life expectancy`)
abline(lsfit(Life_Expectancy_Data$Schooling, Life_Expectancy_Data$`Life expectancy`), lty=2, col='red', lwd=2)
```

# Model Development 
```{r}
m1<-lm(Life_Expectancy_Data$`Life expectancy`~`Adult Mortality`+Alcohol+`percentage expenditure`+`Hepatitis B`+Measles+BMI+`under-five deaths`+Polio+`Total expenditure`+`Diphtheria`+`HIV/AIDS`+GDP+Population+`thinness  1-19 years`+`thinness 5-9 years`+`Income composition of resources`+Schooling, data=Life_Expectancy_Data)
summary(m1)
```

```{r}
vif(m1) #multicolinial check

par(mfrow=c(3,2))
plot(m1)
plot(m1$fitted.values, rstandard(m1),xlab="Fitted values",
ylab="Standardized residuals of model m1")
plot(m1$fitted.values, Life_Expectancy_Data$`Life expectancy`, xlab="m1 fitted values",ylab="1")
abline(lsfit(m1$fitted.values,Life_Expectancy_Data$`Life expectancy`))
```

```{r}
mmp(m1)
```

```{r}
mm1<-lm(Life_Expectancy_Data$`Life expectancy`~`Adult Mortality`+Alcohol+`percentage expenditure`+BMI+`under-five deaths`+`Diphtheria`+`HIV/AIDS`+`Income composition of resources`+Schooling, data=Life_Expectancy_Data)
summary(mm1)
```

```{r}
vif(mm1) #multicolinial check

par(mfrow=c(3,2))
plot(mm1)
plot(mm1$fitted.values, rstandard(mm1),xlab="Fitted values",
ylab="Standardized residuals of model mm1")
plot(mm1$fitted.values, Life_Expectancy_Data$`Life expectancy`, xlab="mm1 fitted values",ylab="1")
abline(lsfit(m1$fitted.values,Life_Expectancy_Data$`Life expectancy`))
```

```{r}
mmp(mm1)
```

```{r}
anova(m1,mm1)
```

### BIC Backward Elimination

```{r}
n=nrow(Life_Expectancy_Data)
backBIC<-step(m1,direction="backward",k=log(n))
```

#### Model developed on BE variable subset
```{r}
m2<-lm(`Life expectancy`~`Adult Mortality`+`percentage expenditure`+BMI+`under-five deaths`+`Diphtheria`+`HIV/AIDS`+`Income composition of resources`+Schooling, data=Life_Expectancy_Data)
summary(m2)
```

```{r}
mmp(m2)
```
```{r}
vif(m2)
```

```{r}
anova(m2,m1) #p-value = 0.1657, Use Transformation

par(mfrow=c(3,2))
plot(m2)
plot(m2$fitted.values, rstandard(m2),xlab="Fitted values",
ylab="Standardized residuals of model m2")
plot(m2$fitted.values, Life_Expectancy_Data$`Life expectancy`, xlab="m2 fitted values",ylab="1")
abline(lsfit(m1$fitted.values,Life_Expectancy_Data$`Life expectancy`))
```

```{r}
plot(m2$fitted.values, rstandard(m2),xlab="Fitted values",ylab="Standardized residuals of m2")
abline(h=4,lty=2,col=6,lwd=2)
abline(h=-4,lty=2,col=6,lwd=2)
#identify(m2$fitted.values,rstandard(m2))
```

#### Power Transform to further refine model
```{r}
summary(powerTransform(cbind(`Life expectancy`,`Adult Mortality`,BMI,`Diphtheria`,`HIV/AIDS`)~1),data=Life_Expectancy_Data)
#Adult - square root
#HIV/AIDS - 1/square root
```

```{r}
tadult<-(Life_Expectancy_Data$`Adult Mortality`)^2
thiv<- 1/sqrt(Life_Expectancy_Data$`HIV/AIDS`)
m3<-lm(`Life expectancy`~tadult+`percentage expenditure`+BMI+`under-five deaths`+`Diphtheria`+thiv+`Income composition of resources`+Schooling, data=Life_Expectancy_Data)
summary(m3)
```

```{r}
par(mfrow=c(2,2))
plot(m3)
```

```{r}
plot(m3$fitted.values, rstandard(m3),xlab="Fitted values",ylab="Standardized residuals of m3")
abline(h=4,lty=2,col=6,lwd=2)
abline(h=-4,lty=2,col=6,lwd=2)
identify(m3$fitted.values,rstandard(m3))
```

```{r}
mmp(m3)
```

```{r}
vif(m3)
```

#### Square root of BIC selected variables
```{r}
m4<-lm(sqrt(`Life expectancy`)~sqrt(`Adult Mortality`)+sqrt(`percentage expenditure`)+sqrt(BMI)+sqrt(`under-five deaths`)+sqrt(`Diphtheria`)+sqrt(`HIV/AIDS`)+sqrt(`Income composition of resources`)+sqrt(Schooling), data=Life_Expectancy_Data)
summary(m4)
```

```{r}
par(mfrow=c(2,2))
plot(m4)
```

```{r}
plot(m4$fitted.values, rstandard(m4),xlab="Fitted values",ylab="Standardized residuals of m4")
abline(h=4,lty=2,col=6,lwd=2)
abline(h=-4,lty=2,col=6,lwd=2)
identify(m4$fitted.values,rstandard(m4))
```

```{r}
mmps(m4)
```
```{r}
vif(m4)
```

```{r}
anova(m2,m4)
```



